---
title: "Volleybuffs_EDA"
format: html
---
#### Loading Libraries
```{r,message =FALSE, warning=FALSE,echo=FALSE}
library(tidyverse)
library(ggplot2)
```


## Bringing in the data

```{r}
#raw_data <- read.csv("~/Documents/Academic/STAT4640_Stats_Capstone/Data/dirty_all_matches_raw.csv")
raw_conf_data <- read.csv("~/Documents/Academic/STAT4640_Stats_Capstone/Data/conference_matches_raw.csv")

# Keep the raw data intact; use this smaller table for metric construction.
core_columns <- c(
  "match_id", "set_number", "point_id", "file_line_number",
  "home_team", "visiting_team", "team",
  "skill", "evaluation_code", "evaluation", "special_code",
  "point", "point_won_by", "winning_attack", "serving_team",
  "home_team_score", "visiting_team_score",
  "home_score_start_of_point", "visiting_score_start_of_point"
)
conference_data <- raw_conf_data |>
  select(all_of(core_columns))
head(conference_data)
```

`conference_data` excludes player names, IDs, jersey numbers, lineup fields,
video/timestamp fields, coordinates, and export metadata. Match, set, rally,
team, and source-line identifiers remain for grouping and checking records;
they are not automatically model predictors. `special_code` remains because
it describes faults such as net touches or balls out and may help us classify
errors. The original CSV and `raw_conf_data` retain every column.

These are inputs for constructing rally-level metrics, not a ready-to-fit
predictor table. `point_won_by` and `winning_attack` describe the completed
rally: use them only after that rally ends. For predictions before a rally,
use prior-rally metrics and the start-of-point scores. The eventual set winner
is the outcome label, never a predictor.


## Separating conference matches by ID

Each entry in `conference_matches` is a data frame containing all rows for one
match. The list is named by `match_id`, and row order within each match is preserved.
There are 104 recorded matches within the conference where all games were home games

```{r}
stopifnot(!anyNA(raw_conf_data$match_id), all(nzchar(raw_conf_data$match_id)))
conference_matches <- split(conference_data, conference_data$match_id)

# A lookup table to identify matches and see how many recorded rows each has.
match_summary <- conference_data |>
  group_by(match_id, home_team, visiting_team) |>
  summarise(recorded_rows = n(), .groups = "drop")

match_summary
length(conference_matches)
```

## Looking at an individual match

Use a match ID from `match_summary` to select a match. These are recorded event
rows, so `recorded_rows` is not a count of points or rallies.

```{r}
#Other ways to select or inspect a match:
first_match <- conference_matches[[1]]
head(first_match)
# View(first_match)  # Open in an interactive R session.
```


## Looking at the just the first match

We're looking for a relationship that would exist between the number of points and the teams chances of winning at that given point.

Split the first match into a named list using `set_number`. Each entry contains
all recorded rows for that set, in their original order. Here, "first match"
means the first entry in `conference_matches`, which is ordered by match ID.

```{r}
stopifnot(!anyNA(first_match$set_number))
first_match_sets <- split(first_match, first_match$set_number)

# See which match and sets we are working with.
unique(first_match[c("match_id", "home_team", "visiting_team")])
set_summary <- first_match |>
  count(set_number, name = "recorded_rows")
set_summary

# Isolate set 1 for our initial analysis.
first_set <- first_match_sets[["1"]]
head(first_set)

# Select another set by its number when needed:
# second_set <- first_match_sets[["2"]]
# View(first_set)  # Open in an interactive R session.
```

## Who won each set, and when?

Use the last point record (`point == TRUE`) in each set to read the final
score. Some `end_of_set` markers are assigned to the next set, so we do not
use those markers to determine the winner. Only label a winner when the
recorded score reaches the set's target with a lead of at least two points.

```{r}
first_match_results <- first_match |>
  mutate(match_event_row = row_number()) |>
  filter(point %in% TRUE) |>
  group_by(match_id, set_number) |>
  slice_tail(n = 1) |>
  ungroup() |>
  mutate(
    target_score = if_else(set_number == 5, 15L, 25L),
    set_complete = !is.na(home_team_score) &
      !is.na(visiting_team_score) &
      pmax(home_team_score, visiting_team_score) >= target_score &
      abs(home_team_score - visiting_team_score) >= 2,
    winning_team = case_when(
      set_complete & home_team_score > visiting_team_score ~ home_team,
      set_complete & visiting_team_score > home_team_score ~ visiting_team,
      TRUE ~ NA_character_
    ),
    total_points_at_end = home_team_score + visiting_team_score
  ) |>
  select(
    match_id, set_number, home_team, visiting_team,
    home_team_score, visiting_team_score, winning_team, set_complete,
    total_points_at_end, point_id, match_event_row, file_line_number
  )

first_match_results

# The result for the set we isolated earlier.
first_set_result <- first_match_results |>
  filter(set_number == 1)
first_set_result
```

`total_points_at_end` tells us how many points had been scored when the set
ended. `point_id` is the raw identifier, not the number of rallies in that set.
`match_event_row` locates the final point record within `first_match`.
The trimmed table tracks when a set ends by point count and event order.
If `set_complete` is false, the last recorded score does not establish a winner.

## Building live performance inputs and initial models

The following script builds one snapshot after each completed rally across all
conference matches. Each snapshot contains cumulative attempts, kills, attack
errors, blocked attacks, aces, service errors, reception quality, recorded errors,
earned scoring actions, sideout rates, and serving-point rates for both teams.
Blocked attacks (`/`) are separate from attacks blocked for reattack (`!`).
Recorded errors are not automatically classified as unforced errors. Earned
scoring actions and errors are event counts, not mutually exclusive point causes.

Raw percentages use a 0–100 scale and are undefined before any attempts.
Attack efficiency uses a -1 to 1 scale. For modeling, rate inputs use fixed
smoothing and home-minus-visiting differences on a proportion scale. All
calculations reset at the start of each set and use only completed rallies.

We retain `winning_attack` for comparison with the Attack/`#` definition. The
`kill_flag_audit` table exposes disagreements for review.

```{r}
#source("R/live_set_models.R")

# Check excluded sets and inconsistent kill flags before interpreting results.
count(set_audit, usable)
kill_flag_audit

first_set_metrics <- rally_metrics |>
  filter(match_id == first_match$match_id[1], set_number == 1)
first_set_metrics |>
  select(rally_number, home_team_score, visiting_team_score,
         home_attack_attempts, home_kills, home_attack_errors,
         home_blocked_attacks, home_kill_pct, home_attack_efficiency,
         away_attack_attempts, away_kills, away_kill_pct) |>
  head(10)
```

One set lets us verify metrics, but training requires multiple set outcomes.
The script excludes sets with incomplete or inconsistent scoring sequences.
It holds out 20% of matches, with every set and rally from a match kept together.
Final set-winning rallies are excluded from model inputs. The eventual winner
is used only as the response label. Each set receives equal total weight.

The first model uses score difference, total points, their interaction, next
serving team, and whether this is a fifth set. The second adds seven skill-rate
differences. Sideout and serving-point metrics remain available for analysis;
we do not add every related metric to the initial model at once.

```{r}
model_comparison
stage_comparison
calibration_table
```

Lower Brier score and log loss indicate better probability predictions.
Compare score-plus-skills with score-only to see whether skill metrics add
information beyond the score. This is an initial holdout comparison, not a
validated coaching rule. Repeated rallies are correlated, so ordinary model
coefficient p-values should not be interpreted as independent-observation
evidence. Subsequent tuning should use grouped cross-validation within training
matches and leave these test matches untouched. Early/middle/late labels here
use total points and should be evaluated separately for shorter fifth sets.

```{r}
# Show prediction updates for a held-out set, rather than a training example.
example_prediction <- test_predictions |>
  filter(match_id == test_ids[1],
         set_number == min(set_number[match_id == test_ids[1]]))

# Use one score record per rally; include the actual final point, but do not
# make a prediction after the outcome is known.
example_score <- rally_metrics |>
  filter(match_id == example_prediction$match_id[1],
         set_number == example_prediction$set_number[1]) |>
  select(rally_number, home_team_score, home_team, visiting_team, target)
score_axis_max <- max(example_score$target, example_score$home_team_score)
score_label <- paste0(example_score$home_team[1], " cumulative score")
example_score <- bind_rows(
  tibble(rally_number = 0L, home_team_score = 0L), example_score
)

prediction_plot <- ggplot(example_prediction,
       aes(x = rally_number, y = win_probability, color = model)) +
  geom_line() +
  geom_step(data = example_score,
            aes(x = rally_number, y = home_team_score / score_axis_max,
                color = "home_score"),
            inherit.aes = FALSE, linewidth = 0.7, linetype = "dashed") +
  scale_color_manual(
    values = c(score_only = "#F8766D", score_plus_skills = "#00BFC4",
               home_score = "#333333"),
    breaks = c("score_only", "score_plus_skills", "home_score"),
    labels = c("Score only", "Score plus skills", score_label)
  ) +
  scale_y_continuous(
    limits = c(0, 1), labels = scales::label_percent(),
    sec.axis = sec_axis(~ . * score_axis_max,
                        name = paste0(example_score$home_team[2], " points"))
  ) +
  labs(x = "Completed rallies", y = "Probability home team wins the set",
       color = NULL, title = "Predictions and home score during a held-out set",
       subtitle = paste(example_prediction$home_team[1], "vs",
                        example_prediction$visiting_team[1], "— set",
                        example_prediction$set_number[1]),
       caption = "Probability: left axis. Dashed home score: right axis. Final point shown for score only.") +
  theme_minimal()
prediction_plot
```

The dashed staircase shows the home team's cumulative points on the right axis.
It rises by one when home wins a rally and stays flat when visiting wins.
The colored lines use the left probability axis. Their vertical intersections
with the score line have no statistical meaning: the axes measure different things.

### What the fitted models use

Both fitted models use logistic regression to predict `home_set_win` (1 for
a home win, 0 for a visiting win). Coefficients are estimated from the training
matches, with equal total weight for each set. The `quasibinomial()` family
uses a logistic probability function and accommodates the fractional weights.

The inputs for the model are used to predict given the current metrics for the  |

| `score_difference` | Home points minus visiting points |
| `total_points` | Home points plus visiting points |
| `score_difference:total_points` | Allows the effect of a lead to change as points accumulate |
| `next_home_serves` | 1 if home will serve the next rally, otherwise 0 |
| `fifth_set` | 1 for the shorter fifth set, otherwise 0 |

The score-plus-skills model adds home-minus-visiting differences in kill rate,
attack-error rate, blocked-attack rate, ace rate, service-error rate,
perfect-reception rate, and reception-error rate. Each rate is cumulative
within the set and smoothed as `(count + 0.5) / (attempts + 1)` for model input.
Displayed raw percentages are unsmoothed. These two models do not yet use
recent rally windows, player identities, PCA, or the other candidate algorithms.

## Candidate models and dimension reduction

These are replacement working snippets for the candidate models discussed in
our notes; the original wording was not recoverable from the workspace or Git.
They are deliberately not executed during rendering. The logistic models above
are initial baselines, not a restriction on which models we will compare.

All candidates should predict home-set win probabilities using the same inputs
and match-grouped splits. Choose tuning parameters using grouped validation
within training matches before evaluating the reserved test matches.

### Common inputs

Standardize continuous inputs using training means and standard deviations.
Keep IDs and the eventual winner out of the predictor matrix. These snippets
use the same score and skill inputs as the initial skill model.

```{r}
#| eval: false
x_train <- model.matrix(skill_formula, train_data)[, -1, drop = FALSE]
x_test <- model.matrix(skill_formula, test_data)[, -1, drop = FALSE]
varying <- apply(x_train, 2, sd) > 0
x_train <- x_train[, varying, drop = FALSE]
x_test <- x_test[, varying, drop = FALSE]
training_center <- colMeans(x_train)
training_scale <- apply(x_train, 2, sd)
x_train_scaled <- scale(x_train, center = training_center, scale = training_scale)
x_test_scaled <- scale(x_test, center = training_center, scale = training_scale)
y_train <- factor(train_data$home_set_win, levels = c(0, 1))
```

### Naive Bayes

A simple probabilistic benchmark. Its conditional-independence assumption may
be restrictive because our skill rates and score measures are related.

```{r}
#| eval: false
nb_fit <- e1071::naiveBayes(x = x_train_scaled, y = y_train)
nb_probability <- predict(nb_fit, x_test_scaled, type = "raw")[, "1"]
```

### K-nearest neighbors (KNN)

Predict using similar rally snapshots. Scaling matters because distance drives
the predictions. Tune the neighbor count within match-grouped training folds;
the value below is only an example. Nearby snapshots from one set can dominate
the neighbors, so consider balanced snapshot sampling per set during training.

```{r}
#| eval: false
knn_fit <- class::knn(train = x_train_scaled, test = x_test_scaled,
                      cl = y_train, k = 25, prob = TRUE)
winning_class_probability <- attr(knn_fit, "prob")
knn_probability <- ifelse(knn_fit == "1", winning_class_probability,
                          1 - winning_class_probability)
```

### Quadratic discriminant analysis (QDA)

Allows a different covariance matrix for each outcome class. Highly correlated
or redundant predictors can cause singular covariance estimates. Reduce the
predictor set or compare PCA inputs if this occurs; check assumptions rather
than treating the large number of rally rows as independent evidence.

```{r}
#| eval: false
qda_fit <- MASS::qda(x = x_train_scaled, grouping = y_train)
qda_probability <- predict(qda_fit, x_test_scaled)$posterior[, "1"]
```

### Neural network

A small network can learn nonlinear relationships. Tune hidden-unit count and
weight decay within training folds. Begin small because the number of distinct
matches and sets is much smaller than the number of rally snapshots.

```{r}
#| eval: false
set.seed(4640)
nn_fit <- nnet::nnet(x = x_train_scaled, y = train_data$home_set_win,
                     size = 3, decay = 0.1, entropy = TRUE,
                     weights = train_data$set_weight,
                     maxit = 500, trace = FALSE)
nn_probability <- as.numeric(predict(nn_fit, x_test_scaled, type = "raw"))
```

### PCA followed by logistic regression

PCA combines correlated predictors into fewer components. It finds directions
of predictor variation, not directions that necessarily predict wins. Compare
its held-out performance with the original predictors; it also makes skill-level
interpretation less direct. PCR conventionally uses those components in linear
regression. For this binary response, use a logistic classifier on the components.

Fit PCA on training data only. Repeat scaling, PCA fitting, and component-count
selection separately inside every validation fold. The three components below
are illustrative, not a tuned choice. This example transforms all inputs; we
could alternatively transform only skill inputs and keep score terms explicit.

```{r}
#| eval: false
pca_fit <- prcomp(x_train_scaled, center = FALSE, scale. = FALSE)
cumulative_variance <- cumsum(pca_fit$sdev^2) / sum(pca_fit$sdev^2)
n_components <- min(3L, ncol(pca_fit$x))
pc_names <- paste0("PC", seq_len(n_components))
pc_train <- as.data.frame(pca_fit$x[, pc_names, drop = FALSE])
pc_test <- as.data.frame(predict(pca_fit, x_test_scaled)[, pc_names, drop = FALSE])
pc_train$home_set_win <- train_data$home_set_win
pc_train$set_weight <- train_data$set_weight
pca_formula <- reformulate(pc_names, response = "home_set_win")
pca_logistic_fit <- glm(pca_formula, data = pc_train,
                         family = quasibinomial(), weights = set_weight)
pca_probability <- predict(pca_logistic_fit, pc_test, type = "response")
```

For each candidate, compare set-weighted Brier score, log loss, calibration, and
performance by set stage. Naive Bayes, KNN, and QDA examples above do not apply
the baseline's set weights during fitting; report that distinction or use a
shared balanced-snapshot training strategy for a fairer comparison. These are
templates until dependencies, grouped tuning, and model diagnostics are checked.
